---
title: "Advanced Computational Systems Biology- scRNAseq"
output: html_notebook
---

# Load libraries and data

Libraries:

```{r}
library(Matrix)
library(Seurat)
library(KneeArrower)
library(tidyverse)
```

Dataset. Starting with just one batch to not to overload our computers:

```{r}
mat <- readMM("C:/Users/Lenovo/Desktop/MolBio/2S/Practicals/Simon/Dataset_reduced/counts_s4d1.mtx") #counts
obs <- read.csv("C:/Users/Lenovo/Desktop/MolBio/2S/Practicals/Simon/Dataset_reduced/obs_s4d1.csv", row.names = 1) #metadata
var <- read.csv("C:/Users/Lenovo/Desktop/MolBio/2S/Practicals/Simon/Dataset_reduced/var.csv", row.names = 1) #genes id 
```

We have to make the files match in info:

```{r}
colnames(mat) <- rownames(var)
rownames(mat) <- rownames(obs)
```


# Data preprocessing

Sparse matrix. 

```{r}
umi.per.cell <- colSums(mat)
umi.per.cell <- sort(umi.per.cell, decreasing = T)
cumulative.sum <- cumsum(umi.per.cell)
plot(cumulative.sum, type = "l")
```

They seem to be already cleaned up.


# Create the Seurat Object

We have to transpose the counts because the stats package is done for reverse position of genes vs samples:

```{r}
seu <- CreateSeuratObject(
  t(mat),
  meta.data = obs[,c("cell_type", "batch", "Samplename", "DonorNumber")]
)
```

## Further QC metrics

```{r}
VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA"), ncol = 3, pt.size = 0.1)
```

```{r}
summary(seu$nFeature_RNA)
summary(seu$nCount_RNA)
```

They also seem to be processed in this way.

# Data analysis

## Normalization

```{r}
seu <- NormalizeData(seu)
```


## Find Variable Features

```{r}
seu <- FindVariableFeatures(seu)
```


## Scaling

```{r}
seu <- ScaleData(seu)
```


## PCA 

Linear dimensionality reduction --> weighted sum --> linear combination of the previous dimensions (all genes) to describe more easily our data variability.

```{r}
seu <- RunPCA(seu, ndims = 50)
```


```{r}
ElbowPlot(seu, ndims = 50)
```


## UMAP

Non-linear dimensionality reduction. Keep the 20 first dimensions:

```{r}
seu <- RunUMAP(seu, dims = 1:20)
```


### Plotting

UMAPPlot function or FeaturePlot 

```{r}
UMAPPlot(seu) + coord_equal()
```


## Clustering

### Find neighbors

```{r}
seu <- FindNeighbors(seu, dims = 1:20)
```


### Find clusters

```{r}
seu <- FindClusters(seu, resolution = c(0.1, 0.5, 0.7, 1, 1.5))
```

#### Plotting

```{r}
SetIdent(seu, value = "orig.ident")
DimPlot(seu, group.by = "RNA_snn_res.0.1", label = TRUE)
```

```{r}
DimPlot(seu, group.by = "RNA_snn_res.0.5", label = TRUE)
```
```{r}
DimPlot(seu, group.by = "RNA_snn_res.1.5", label = TRUE)
```

### See the difference with cell types (cheat a bit hehe)

```{r}
DimPlot(seu, group.by = "cell_type", label = TRUE)
```

```{r}
FindMarkers(seu, ident.1 = c(0,1,4))
```

## Anotating cell types



```{r}
FeaturePlot(seu, "CD3E")
```


```{r}
seu@meta.data
```

Show the counts of a housekeeping gene:

```{r}
seu@assays$RNA@counts["ACTB", 1:30]
```

### Double check the normalization

Equation: 

y = log (k/s·10^4 + 1) / 
  k = seu@assays$RNA@counts(counts)
  y = seu@assays$RNA@data (output)
  s = seu@meta.data$nCount_RNA 

```{r}
plot(colSums(seu@assays$RNA@counts), seu@meta.data$nCount_RNA)
```

This line indicates that is ok (?)

And for the housekeeping gene. Normalize it applying the previous equation with the housekeeping gene expression to see if it worked. We add (+1) inside the log so that, if a count is 0 (what is very often), we get log(1)=0 and not log(0)=-inf.

```{r}
plot(log(seu@assays$RNA@counts["ACTB",]/seu@meta.data$nCount_RNA * 1e4 + 1), seu@assays$RNA@counts["ACTB",])
```

But this +1 add a big bias to the results, mainly separating cells that express and do not express it, and for genes with lower counts (?).

See the counts in histogram. (!) I DID NOT FOLLOW THAT PART AND COULD NOT COPY IT ALL CORRECTLY (!): 

```{r}
hist(log10(seu@meta.data$nCount_RNA), 100) #weird done

hist(seu@assays$RNA@counts["ACTB",]) #weird done

plot(seu@meta.data$nCount_RNA, seu@assays$RNA@data["ACTB",], + rnorm(nrow(mat), 0, 0.1), log = "x", pch = ".")
```

### What is the average expression (on the log scale) of ACTB in the cells of Cluster 3 ?

Set the seed first for us all to have the same results (and rerun the analysis quickly):

```{r}
set.seed(12345678)
seu <- RunUMAP(seu, dims = 1:20)
seu <- FindNeighbors(seu, dims = 1:20)
seu <- FindClusters(seu, resolution = 0.5)

UMAPPlot(seu, label=T)
```

Which genes are in cluster 3 and get the mean for ACTB gene:

```{r}
logic_cluster3 <- seu@meta.data$seurat_clusters == 3 #Gives back a logic matrix with T for all cells that have been clustered to cluster 3 by UMAP --> can be used afterwards for subsetting

#seu@assays$RNA@data["ACTB",logic_cluster3]
# With that line, we get the wor for ACTB and all columns from cluster 3

mean(seu@assays$RNA@data["ACTB",logic_cluster3]) #Mean of all columns for cluster 3 for row "gene ACTB" 
```

Now we get the mean of all rows(all genes), but only the column for genes in cluster 3:

```{r}
rowMeans(seu@assays$RNA@data[, logic_cluster3]) #mean gene expression of all genes for cluster 3 
```

A function to get the fene means for a cluster:

```{r}
gene_means_for_cluster <- function(cl){
  w <- seu@meta.data$seurat_clusters == cl
  rowMeans(seu@assays$RNA@data[,w])
}
```

Now we can call it for all clusters for a gene of interest:

```{r}
gene_means_for_cluster(3)["ACTB"]
```

Do it for all clusters. For that, we use sapply, which calls the function for all clusters. It saves it in a matrix, getting the average gene expression of each gene for each cluster:

```{r}
gene_means <- sapply(0:14, gene_means_for_cluster)
gene_means[1:5, 1:5]
```

Now we can extract directly from this matrix the mean gene expression for ACTB in cluster 3:

```{r}
gene_means["ACTB",4] #VERY IMPORTANT, WE HAVE TO SELECT THE COLUMN 4 BECAUSE THE INDEXING OF COLUMNS START AT 1 BUT OUR FIRST CLUSTER IS 0, SO CLUSTER 3 CORRESPONDS TO COLUMN 4
```
(!) VERY IMPORTANT THE NOTE OF THE PREVIOUS CHUNK (!)

Consequently, we have to change the col names beforehand:

```{r}
colnames(gene_means) = sprintf("CL_%02d", 0:14) #function string print formatting for putting the number of the clusters (not need to understand)

gene_means[1:5, 1:5]
```

So now we can:

```{r}
gene_means["ACTB","CL_03"]
```

```{r}
gene_means["ACTB",4] == gene_means["ACTB","CL_03"]
```

## Find genes that differ a lot from cluster to cluster

Calculate the SD/variance of the mean matrix:

```{r}
library(matrixStats)

gene_vars_accorss_clusters <- rowVars(gene_means, useNames = T) #useNames to not to lose them after the analysis
```

Now we want to keep the top 20 most differentially expressed genes:

```{r}
interesting_genes <- names(sort(gene_vars_accorss_clusters, decreasing = T)[1:20])
```

Subset the matrix with gene means:

```{r}
heatmap(gene_means[interesting_genes,])
```

We see our sample separated in 4 clusters.

For 100 genes:

```{r}
interesting_genes <- names(sort(gene_vars_accorss_clusters, decreasing = T)[1:50])

heatmap(gene_means[interesting_genes,])
```


```{r}
#centered_gene_means <- rowMeans(gene_means[interesting_genes,] - rowMeans(gene_means[interesting_genes,]))
centered_gene_means <- gene_means - rowMeans(gene_means)

heatmap(centered_gene_means[interesting_genes,], 
        scale = "none", 
        zlim = c(-4,4), 
        col = colorRampPalette(c("blue", "grey", "red"))(100)) #Red for high, blue for low
```

We can see that clusters 5, 6 and 14 share some profile --> CD36 --> look for literature to see which follow the profile. 

```{r}
UMAPPlot(seu, label =T)
```

We can guess that 5 and 6 are the progenitors and the 14 are the mature ones that are being pushed out from the bone marrow (this needs further study).

We can also ask ChatGPT (and contrast afterwards, of course), e.g., give me gene markerts that separate the myeloid and lymphoid lines in bone marrow. Create a vector with our interesting genes and test them:

```{r}
lymphoid_genes <- c("CD19", "CD3G", "CD3E", "MS4A1", "PAX5", "CD79B")
```

Clusters 3,7,9,10 --> LYMPHOID
Clusters

Do same with myeloid.

And we can also look for individual genes in plots:

```{r}
FeaturePlot(seu, "CD33", label = T)
```

It is expressed in cluster 2, 12, etc. but still not very expressed --> not very well seen in the Feature plot but better with means or violin plot:

```{r}
VlnPlot(seu, "CD33") 
```

We do not see much because it is a not very expressed gene, but it is important. Maybe we could remove the 0 to see the actual distribution (not done here):

Here with means. It is expressed in few cells but very high so increases the mean:

```{r}
plot(gene_means["CD33",])
```

We can try with other gene to see a better example:

```{r}
FeaturePlot(seu, "PAX5", label = T)
VlnPlot(seu, "PAX5")
plot(gene_means["PAX5",])
```
Caracteristic of clusters 3,7,9,10 --> from the lymphoid line.

This analysis lasts several days to properly anotate cell types. Checking again against the real anotations of the dataset:

```{r}
UMAPPlot(seu, group.by = "cell_type", label = T)
```

To interact with the UMAP space --> we can see the direct relation (distances) of each cell with the surrounding ones: 

```{r}
sleepwalk::sleepwalk(seu@reductions$umap@cell.embeddings, seu@reductions$pca@cell.embeddings)
```

Batch effect:

What is larger in our sample?
- Two different cells types of the same donor
- Same cell type of different donor (this should differ less) --> But also depending on the donors (samples) --> healthy vs sick

Clusters with different subgroups can be the different patients, or activated/deactivated immune cells, progenitors and intermediate developmental states, etc.

When the patients are more different (batch effect), the clusters are more elongated in shape and the algorithm can identify them as different clusters.

Integration methods. 



Are clusters representing real cell types or different donors/batches? --> look for batch effect. 

We can group.by = "batch" and compare with the clusters obtained before. Maybe in different clusters, the batch representation can be even or the cluster can represent each different batch --> it has to be corrected. 

After properly labelling, we can see the differentiation (development) of bone marrow cells. 

Which genes are absolutely representative and which ones change from person to person?


# Compare different clusters

Compare the T cell CD8 and CD4 (both naive and activated) --> corresponding to clusters 4 and 1 respectively.



```{r}
CD8_cells <- seu@meta.data$cell_type == "CD8+ T" #I could have used seu@meta.data$seurat_clusters == 4 (and 1 for CD4)
CD4_cells <- seu@meta.data$cell_type == c("CD4+ T activated", "CD4+ T naive")
```


```{r}
mean_gene_exp_CD8 <- rowMeans(seu@assays$RNA@data[,CD8_cells])
mean_gene_exp_CD4 <- rowMeans(seu@assays$RNA@data[,CD4_cells])
```

```{r}
plot(mean_gene_exp_CD4, mean_gene_exp_CD8, asp =1) 
```

Another way of plotting it. It is how far I am from the diagonal (diagonal above as x axis). Should be the same plot as before:

```{r}
plot(mean_gene_exp_CD4+mean_gene_exp_CD8/2, mean_gene_exp_CD4 - mean_gene_exp_CD8, cex = 0.3)
```

Poisson noise --> That is why it gets broader in the middle and then gets thinner again. This effect is greater in low expressed genes. 

```{r}
identify(plot(mean_gene_exp_CD4+mean_gene_exp_CD8/2, mean_gene_exp_CD4 - mean_gene_exp_CD8, cex = 0.3)) #I don't get wtf is Identify
```

Sort the difference in mean expression among them and see the ones with higher for CD8 (tail, lower difference when substracting CD4 means --> pos values) and higher differences for CD4 (head, higher difference when substracting CD4 means --> neg):

```{r}
tail(sort (mean_gene_exp_CD8 - mean_gene_exp_CD4))
```

```{r}
head(sort (mean_gene_exp_CD8 - mean_gene_exp_CD4))
```

The gene AOAH is very different for CD8 --> very interesting --> we can google it.
INPP4B --> very charasterstic in CD4.

```{r}
plot(mean_gene_exp_CD8 - mean_gene_exp_CD4, sqrt(var(mean_gene_exp_CD4 + mean_gene_exp_CD8)/2), cex = 0.3) #I didn't get it and it doesn't work for me
```

```{r}
t.test(
  seu@assays$RNA@data[,CD8_cells],
  seu@assays$RNA@data[,CD4_cells]
)
```

We have seen significance but it does not mean much because they were clusters already (?).


```{r}
library(tidyverse)

tibble(
  cell = rownames(mat),
  gene_count = mat[,"AOAH"],
  total = rowSums(mat),
  cell_type = obs$cell_type,
  batch = obs$batch,
  donor = obs$DonorNumber) -> tbl
```


```{r}
tbl %>%
  filter(cell_type %in% c("CD4+ T naive", "CD4+ T activated", "CD8+ T")) %>%
  mutate(simple_cell_type = str_sub(cell_type, 1, 6)) %>%
  mutate(expr = log(gene_count/total * 1e4 + 1)) %>%
  group_by(simple_cell_type, donor) %>%
  summarise(mean(expr), n()) 

pivot_wider(names_from = "simple_cell_type", values_from = "mean(expr)") -> tbl2 #this code line doesn't work, I got lost
t.test(tbl2$'CD4 +', tbl2$'CD8 +', paired=T) #doesn't work either
```


# ADDITIONAL NOTES

## Poisson

Taking balls from a bag and seeing the number of red balls --> taking mRNA from a cell and seeing which is my mRNA of interest.
Poisson is an approximation of a Binomial distribution.

k ~ Pois (lambda)

E(k) = lambda; Var(k) = lambda
SD (k/lambda) = sqr(Var(k/lambda)) = sqr(1/lambda^2[Var(k)]) = 1/sqr(lambda)

When there is a big difference in a given gene expression among cells, the 1st question we have to ask ourselved is if that is real or if that is because the gene is very low expressed --> variance (Poisson) is higher if the gene expression is low than if it is high (the variation is lowered at high counts). Poisson noise!

## Notation in R

For complex objects:

@ for selecting matrices (dfs) in the object.
$ for selecting columns in matrices.

Basically, we have to alter them and it works

## Taking an average before and after a log transformation

log transformation is a way of getting the number of digits i.e. the order of the numbers.

It is more representative after log. Example:

9x 1000$
1x 1000000$

Normal mean income: 100900$

Mean after log and then reverse log transform: 
  9x 3
  1x 6
  Mean: 3.3 --> rev log(10) transform --> 2000$ --> more representative
  
  
# Save RDS

```{r}
saveRDS(seu, "seu_BM_s4d1_1.rds")
```

```{r}
sessionInfo()
```



